---
title: "Confusion:MaxEnt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
layout: post
# knit: (function(inputFile, encoding){
#        rmarkdown::render(inputFile,
#        encoding=encoding)})
categories:
  - Miscellaneous
tags:
  - content
output:
  md_document:
    preserve_yaml: TRUE
    variant: markdown_strict 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
